Optimizing Abstract Abstract Machines

### 2012-11-14 paper/paper.tex
Cleaned up time-stamping the store

\begin{align*}
\force &: \Store \times \Addr \to \Set(\Value) \\
\force(\msto,\saddr\maddralt) &= \msto(\maddralt)\\
\force(\msto,\mval) &= \{\mval\} \\
\ext(\msto, \maddr, \mval) &= \msto \sqcup[\maddr \mapsto \force(\msto, \mval)] \\
\ev{\svar{\mvar},\menv,\mkont,\msto} &\machstep\;
\co{\mkont,\saddr{\menv(\mvar)},\msto} \\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto}
&\machstep\;
\ev[^\mcntr]{\mexp,\menv,\ext(\msto, \maddr^f,\mval),\kfn[^\mcntr_\mlab]{\maddr^f,\maddr}} \\
\mathit{bind\,}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
  \maddr], \ext(\msto,\maddr, \mval), \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= \mvar\mcntr'
\end{align*}

@@

The laziness optimization saved time by doing the same amount of
reasoning as before but in fewer transitions. We can exploit the same
idea - same reasoning, fewer transitions - with abstract
compilation. Abstract compilation is \emph{precision-preserving} and
transforms complex expressions whose \emph{abstract} evaluation is
deterministic into ``abstract bytecodes.''  The abstract interpreter
then does in one transition what previously took many. Abstract
compilation eliminates unnecessary allocation, deallocation and
branching.

The essence of the compilation effect can be seen by considering an
example such as
\[
\sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3}
\]
which makes the following transitions:
\begin{align}
& \ev{\sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3},\menv,\mkont,\msto_0}\\
\machstep\; &
\ev{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2},\menv,\kar{\mexp_3,\menv,\maddr_1},\msto_1}
\\
\machstep\; &
\ev{\sapp\mvar{\mexp_1},\menv,\kar{\mexp_2,\menv,\maddr_2},\msto_2}
\\
\machstep\; &
\ev{\mvar, \menv,\kar{\mexp_1,\menv,\maddr_3},\msto_3} % {\mexp_2}
\\
\machstep\; &
\co{\kar{\mexp_1,\menv},\mval,\msto_4} % {\mexp_1}{\mexp_2}
\mbox{ where } \mval \in \msto(\menv(\maddr))
\end{align}

@@

Every step the analysis makes for the above techniques requires
joining large stores together. Not every step will modify all
addresses of the store, so joining entire stores is wasteful in terms
of memory and time. We can instead log store changes and replay the
change log on the full store after all steps have completed. This uses
far fewer join operations, leading to less overhead, and is \emph{precision preserving}.

We represent changes as $\msdiff \in \StoreDelta = List[\Addr \times
  \Set(\Storeable)]$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$
instead becomes a log addition
$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins
empty ($\mtlst$) for each step. Applying the changes to the full store
is straightforward:

\begin{align*}
\replay &: \StoreDelta \times \Store \to \Store \\
\replay(\mtlst, \msto) &= \msto \\
\replay(\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}, \msto) &= \replay(\msdiff, \msto\sqcup[\maddr \mapsto \mval{s}])
\end{align*}

The reduction relation does have to change to accommodate this
technique, but only in this regard. The whole store is still given so
that lookups are possible. Example rule:

@@

Here $\concat : \Set(\StoreDelta) \to \StoreDelta$ flattens the lists
of changes to one; order it appends does not matter:
\begin{align*}
\concat(\varnothing) &= \mtlst \\
\concat(\{\msdiff\}\cup D) &= \append(\msdiff, \concat(D))
\end{align*}

@@

And, there is a better way. Shivers' original work on k-CFA was
susceptible to the same problem, and he suggested three complementary
optimizations: (1) make the store global; (2) update the store
imperatively; and (3) associate every change in the store with a
version number -- its timestamp. Then, put timestamps in states
where previously there were stores. Given two states, the analysis can
now compare their stores just by comparing their timestamps -- a
constant-time operation.

There is a subtle loss of precision and Shivers' original time-stamp
technique that we can fix. For a given abstract state, all writes to
the global store need to be delayed until the analysis considers all
branches from that abstract state. This avoids cross-branch pollution
which would not otherwise happen, e.g., when one branch writes to
address $\maddr$ and another branch reads from address
$\maddr$. Fortunately, given our conversion to log-based stores this
change is straightforward.

The current state of the state-space search now has six components:

\begin{itemize}
 \item{$\msto$, the store}
 \item{{\tt todo}, the workset}
 \item{{\tt seen}, a map from states to the timestamps at which they were last seen}
 \item{$\msdiff$, the store changes for the current step}
 \item{$\changep$, a boolean tracking whether the stepped state contributed a store change}
 \item{{\tt T}, the timestamp of the store}
\end{itemize}

The new $eval$ calculation is defined in
figure~\ref{fig:imperative}.  To ensure termination, we guard adding
$c$ to {\tt todo} by the following check:
\begin{align*}\changep \vee {\tt seen}(c) \neq {\tt T}\end{align*}
If it succeeds, {\tt todo} gets $c$ and we update {\tt seen} to map $c$ to {\tt T}.

@@

Practical issues of performance arise when one chooses the
``obvious,'' or ``natural'' abstraction for compound data. Our running
example will be the classic Boyer benchmark. The detail of note for
the Boyer benchmark is that it has large list literals, the exact
values of which do not matter for many analyses. If we
interpret a list literal of 30 elements (that again have 10 elements
each) as 300 applications of {\tt cons}, and then interpret a function
that traverses the tree, we will end up interpreting that function at
least 300 times, when once or twice would do.

>>>>

\begin{align*}
\force &: \Store \times \Value \to \Set(\Value) \\
\force(\msto,\saddr\maddralt) &= \msto(\maddralt)\\
\force(\msto,\mval) &= \{\mval\} \\
\ext(\msto, \maddr, \mval) &= \msto \sqcup[\maddr \mapsto \force(\msto, \mval)] \\
\ev{\svar{\mvar},\menv,\mkont,\msto} &\machstep\;
\co{\mkont,\saddr{\menv(\mvar)},\msto} \\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto}
&\machstep\;
\ev[^\mcntr]{\mexp,\menv,\ext(\msto, \maddr^f,\mval),\kfn[^\mcntr_\mlab]{\maddr^f,\maddr}} \\
\mathit{bind\,}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
  \maddr], \ext(\msto,\maddr, \mval), \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= \mvar\mcntr'
\end{align*}

@@

The prior optimization saved time by doing the same amount of
reasoning as before but in fewer transitions. We can exploit the same
idea---same reasoning, fewer transitions---with abstract
compilation. Abstract compilation is \emph{precision-preserving} and
transforms complex expressions whose \emph{abstract} evaluation is
deterministic into ``abstract bytecodes.''  The abstract interpreter
then does in one transition what previously took many. In short, abstract
compilation eliminates unnecessary allocation, deallocation and
branching.

The following example illustrates
the essence of abstract compilation effect:
\[
\sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3}
\]
makes the following transitions:
\begin{align}
& \ev{\sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3},\menv,\mkont,\msto_0}\\
\machstep\; &
\ev{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2},\menv,\kar{\mexp_3,\menv,\maddr_1},\msto_1}
\\
\machstep\; &
\ev{\sapp\mvar{\mexp_1},\menv,\kar{\mexp_2,\menv,\maddr_2},\msto_2}
\\
\machstep\; &
\ev{\mvar, \menv,\kar{\mexp_1,\menv,\maddr_3},\msto_3} % {\mexp_2}
\\
\machstep\; &
\co{\kar{\mexp_1,\menv},\mval,\msto_4} % {\mexp_1}{\mexp_2}
\mbox{ where } \mval \in \msto(\menv(\maddr))
\end{align}

@@

Every step the analysis makes for the above techniques requires
joining large stores together. Not every step will modify all
addresses of the store, so joining entire stores is wasteful in terms
of memory and time. We can instead log store changes and replay the
change log on the full store after all steps have completed. This uses
far fewer join operations, leading to less overhead, and is \emph{precision-preserving}.

We represent change logs as $\msdiff \in \StoreDelta = (\Addr \times
  \Set(\Storeable))^*$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$
 becomes a log addition
$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins
empty ($\mtlst$) for each step. Applying the changes to the full store
is straightforward:
% TODO: Turn the recursion here into ...:
\begin{align*}
\replay &: \StoreDelta \times \Store \to \Store \\
\replay(\mtlst, \msto) &= \msto \\
\replay(\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}, \msto) &= \replay(\msdiff, \msto\sqcup[\maddr \mapsto \mval{s}])
\end{align*}

The transition relation is identical except for the addition of this change log.
%
Lookups will never rely on the change log, and can use the originally supplied store unmodified:

@@

Here $\concat : \Set(\StoreDelta) \to \StoreDelta$ flattens the lists
of changes to one; the order in which it appends does not matter:
\begin{align*}
\concat(\varnothing) &= \mtlst \\
\concat(\{\msdiff\}\cup D) &= \append(\msdiff, \concat(D))
\end{align*}

@@

And, there is a better way. Shivers' original work on $k$-CFA was
susceptible to the same problem, and he suggested three complementary
optimizations: (1) make the store global; (2) update the store
imperatively; and (3) associate every change in the store with a
version number -- its timestamp. Then, put timestamps in states
where previously there were stores. Given two states, the analysis can
now compare their stores just by comparing their timestamps -- a
constant-time operation.

There is a subtle loss of precision in Shivers' original time-stamp
technique that we can fix. For a given abstract state, all writes to
the global store need to be delayed until the analysis considers all
branches from that abstract state. This avoids cross-branch pollution
which would not otherwise happen, e.g., when one branch writes to
address $\maddr$ and another branch reads from address
$\maddr$. Fortunately, given our conversion to log-based stores this
change is straightforward.

At this point, we can begin to think of the analysis as an imperative
algorithm with six major components:
%
\begin{itemize}
 \item{$\msto$, the store;}
 \item{{\tt todo}, the workset;}
 \item{{\tt seen}, a map from states to the timestamps at which they were last seen;}
 \item{$\msdiff$, the store changes for the current step;}
 \item{$\changep$, a boolean tracking whether the stepped state contributed a store change; and}
 \item{{\tt T}, the timestamp of the store.}
\end{itemize}

The new $\eval$ calculation is defined in
figure~\ref{fig:imperative}.  To ensure termination, we guard against adding
$c$ to {\tt todo} by the following check:
\begin{align*}\changep \vee {\tt seen}(c) \neq {\tt T}\end{align*}
If it succeeds, {\tt todo} gets $c$ and we update {\tt seen} to map $c$ to {\tt T}.

@@

Performance issues arise when one chooses the
``obvious,'' or ``natural'' abstraction for compound data structures like list. 
%
Our running
example will be the classic Boyer benchmark. The detail of note for
the Boyer benchmark is that it has large list literals, the exact
values of which do not matter for many analyses. If we
interpret a list literal of 30 elements (that again have 10 elements
each) as 300 applications of {\tt cons}, and then interpret a function
that traverses the tree, we will end up interpreting that function at
least 300 times, when once or twice would do.
